Skew-symmetric entropy stable modal discontinuous Galerkin formulations

### 2019-03-07 docs/skewform/arxiv/main.tex

\pdfoutput=1

\documentclass{svjour3}                     % onecolumn (standard format)
%\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
%\documentclass[smallextended]{svjour3}       % onecolumn (second format)
%\documentclass[twocolumn]{svjour3}          % twocolumn

@@

\makeatletter
\renewcommand*{\@thmcounterend}{}
\makeatother
%\spnewtheorem{lemma}{Lemma}{\bfseries}{\itshape}
\spnewtheorem{assumption}{Assumption}{\bfseries}{\itshape}
%\spnewtheorem{remark}{Remark}{\bfseries}{\itshape}
\usepackage{cite}

\usepackage{slashbox} % for slashbox in tables

\usepackage{amsmath, amssymb, amsfonts}
%\newtheorem*{remark}{Remark}
%\theoremstyle{assumption}
%\newtheorem{assumption}{Assumption}
\usepackage{hyperref}
\usepackage[titletoc,toc,title]{appendix}
\usepackage{tikz}
\usepackage{hhline}
\usepackage{graphicx, color, bm}
\usepackage{subfig}
\usepackage{pgfplots}
\usepackage{pgfplotstable}
\usepackage{slopetri} % my own package for drawing triangle slopes

@@

\date{}
\author{Jesse Chan}
\title{Skew-symmetric entropy stable discontinuous Galerkin formulations with applications to hybrid meshes}
\titlerunning{Skew-symmetric entropy stable DG formulations}

@@

\begin{abstract}
Entropy stable high order methods for nonlinear conservation laws satisfy an inherent discrete entropy inequality.  The construction of such schemes has relied on the use of carefully chosen collocation points \cite{gassner2013skew, fisher2013high, carpenter2014entropy, crean2018entropy, chan2018efficient} or volume and surface quadrature rules \cite{chan2017discretely, chan2018discretely} to produce operators which satisfy a summation-by-parts (SBP) property.  In this work, we show how to construct skew-symmetric schemes which are entropy stable even for volume and surface quadratures under which a traditional SBP property does not hold.  These skew-symmetric formulations avoid the use of a ``strong'' SBP property, and require only that operators exactly differentiate constants and satisfy a discrete form of the fundamental theorem of calculus.   We conclude with an application of the skew-symmetric formulation for entropy stable schemes on mixed quadrilateral-triangle meshes. 
\end{abstract}

\tableofcontents

@@

These issues have motivated the introduction of high order \textit{entropy stable} discretizations, which satisfy a semi-discrete entropy inequality while maintaining high order accuracy in smooth regions.  Proofs of continuous entropy inequalities rely on the chain rule, which does not hold discretely due to effects such as quadrature error.   By using an approach referred to as ``flux differencing'', entropy stable schemes account for the loss of the chain rule while maintaining a semi-discrete analogue of the continuous entropy inequality.  These schemes were first introduced as high order collocation methods on tensor product elements in \cite{fisher2013high, carpenter2014entropy, gassner2016split, gassner2017br1}, and were extended to simplicial elements in \cite{crean2017high, chen2017entropy, crean2018entropy, chan2017discretely, chan2018discretely}.  Entropy stable methods have also been extended to a variety of other discretization settings, including staggered grids \cite{parsani2016entropy}, Gauss-Legendre collocation \cite{chan2018efficient}, and non-conforming meshes \cite{friedrich2017entropy}.

Entropy stable discretizations are built upon flux differencing and a summation-by-parts (SBP) property, which serves as a discrete analogue of integration by parts for quadrature-based discretization matrices.  However, the SBP property does not hold for some under-integrated quadrature rules, which naturally arise in certain discretization settings.  For example, on hybrid meshes consisting of both quadrilateral and triangular elements, it is convenient to utilize the same quadrature rule on shared faces between different element types.  On degree $N$ tensor product elements, a popular choice of quadrature is an $(N+1)$-point Gauss-Legendre-Lobatto (GLL) rule.  When both volume and surface integrals are approximated using $(N+1)$ point GLL quadrature rules, the SBP property holds, despite the fact that GLL quadrature is inexact for the integrands which appear in finite element formulations \cite{fisher2013high}.  However, while GLL quadrature induces an SBP property on quadrilateral elements, but does not always result in an SBP property if used for triangular elements \cite{chan2017discretely}.  %A similar issue arises on non-conforming meshes, where a composite ``split-side'' quadrature naturally when coupling larger elements to smaller elements through a non-conforming interface \cite{kozdon2018energy}.  For quadrilaterals utilizing a GLL volume quadrature, the use of a composite quadrature on a non-conforming face also results in the loss of the SBP property \cite{chan2017discretely}.

@@

It should be noted that a similar approach to entropy stable discretizations was introduced within a finite difference framework \cite{chen2017entropy, crean2018entropy} using multidimensional differencing operators which are constructed to satisfy accuracy conditions and an SBP property \cite{hicken2016multidimensional}.  These operators do not correspond to any specific basis or approximation space, but can be shown to exist for nodal points corresponding to sufficiently accurate choices of volume and surface quadrature and can be computed either algebraically or through an optimization problem.  The formulations presented in this work require similar accuracy conditions, but differ from those of \cite{chen2017entropy, crean2018entropy} in that we start with an explicit approximation space, which then induce quadrature-based operators.  The formulations presented here also accomodate general choices of volume and surface quadrature (e.g.\ volume quadratures without boundary nodes and over-integrated rules) while ensuring compact coupling conditions between neighboring elements.

The structure of the paper is as follows: Section~\ref{sec:nonlin} describes the continuous entropy inequality which we aim to replicate discretely.  Section~\ref{sec:approx} and Section~\ref{sec:sbp} introduce polynomial approximation spaces and quadrature-based SBP operators on simplicial and tensor product elements.  Section~\ref{sec:skew1} then introduces the variational SBP property and an entropy conservative skew-symmetric formulation.  Section~\ref{sec:skew2} extends the skew-symmetric formulation to curved elements, and Section~\ref{sec:num} presents numerical experiments which verify the theoretical assumptions, stability, and accuracy of the proposed formulations.

@@

The proof of (\ref{eq:entropyineq}) requires the use of the chain rule \cite{mock1980systems, harten1983symmetric, dafermos2005compensated}.  The instability-in-practice of high order schemes for (\ref{eq:nonlineqs}) can be attributed in part to the fact that the discrete form of the equations do not satisfy the chain rule, and thus do not satisfy (\ref{eq:entropyineq}).  As a result, discretizations of (\ref{eq:nonlineqs}) do not typically possess an underlying statement of stability.  For low order schemes, this can be offset in practice by the inherent numerical dissipation.  However, because high order discretizations possess low numerical dissipation, the lack of an underlying discrete stability has been conflated with the idea that high order methods are inherently less stable than low order methods.

@@

In this work, we consider either simplicial reference elements (triangles and tetrahedra) or tensor product reference elements (quadrilaterals and hexahedra).  We define an approximation space using degree $N$ polynomials on the reference element; however, the natural polynomial approximation space differs depending on the element type \cite{chan2015gpu}.

On a $d$-dimensional reference simplex, the natural polynomial space are total degree $N$ polynomials 
\[
P^N\LRp{\hat{D}} = \LRc{\hat{x}_1^{i_1}\ldots\hat{x}_d^{i_d}, \quad \hat{\bm{x}} \in \hat{D}, \quad 0\leq \sum_{k=1}^d i_k \leq N}.
\]
In contrast, the natural polynomial space on a $d$-dimensional tensor product element is the space of maximum degree $N$ polynomials
\[
Q^N\LRp{\hat{D}} = \LRc{\hat{x}_1^{i_1}\ldots\hat{x}_d^{i_d}, \quad \hat{\bm{x}} \in \hat{D}, \quad 0\leq i_k \leq N, \quad k = 1,\ldots, d}.
\]
We denote the natural approximation space on a given reference element $\hat{D}$ by $V^N$.  Furthermore, we denote the dimension of $V^N$ as $N_p = {\rm dim}\LRp{V^N\LRp{\hat{D}}}$.

The proofs presented in this work will also use anisotropic tensor product polynomial spaces, where the maximum polynomial degree varies depending on the coordinate direction.  We denote such spaces by $Q^{N_1, \ldots, N_d}$, where $N_k$ are non-negative integers and
\[
Q^{N_1, N_2, \ldots, N_d}\LRp{\hat{D}} = \LRc{\hat{x}_1^{i_1}\ldots\hat{x}_d^{i_d}, \quad \hat{\bm{x}} \in \hat{D}, \quad 0\leq i_k \leq N_k, \quad k = 1,\ldots, d}.
\]
For example, the isotropic tensor product space $Q^N$ is the same as $Q^{N,\ldots,N}$.

@@

It was shown in \cite{chan2017discretely, chan2018efficient} that, when combined with projection and lifting matrices, $\bm{D}^i_N$ produces a high order polynomial approximation of $f\pd{g}{x_i}$. 
Let $f, g$ be differentiable functions, and let $\bm{f}_i = f(\bm{x}_i)$, $\bm{g}_i = g(\bm{x}_i)$ denote values of $f,g$ at volume and surface quadrature points.  Then,
\begin{gather}
f\pd{g}{x_i} \approx \LRs{\begin{array}{cc}
\bm{P}_q & \bm{L}_f\end{array}} {\rm diag}\LRp{\bm{f}}\bm{D}^i_N \bm{g}.  
\label{eq:dnapprox}
\end{gather}
%The approximation can also be interpreted as solving a variational problem.  Let $n_i$ be the $i$th component of the unit normal on $\hat{D}$.  Then (\ref{eq:dnapprox}) is equivalent to finding $u(\bm{x}) \approx f\pd{g}{x_i}$ such that, for all $v\in V^N\LRp{\hat{D}}$, 
%\begin{align*}
%\int_{\hat{D}} u(\bm{x})v(\bm{x}) = \int_{\hat{D}}{g\pd{\Pi_Nf}{x_i}v} + \int_{\partial \hat{D}}{(f-\Pi_Nf)\frac{\LRp{gv + \Pi_N(gv)}}{2}}\hat{n}_i\hat{J}_f.
%\end{align*}

For sufficiently accurate quadrature rules, the matrix $\bm{Q}^i_N$ also satisfies a decoupled summation-by-parts (SBP) property, which is used to prove semi-discrete entropy stability for nonlinear conservation laws.  
\begin{theorem}%[Theorem 1 in \cite{chan2017discretely}]
Assume that $\bm{Q}^i$ satisfies a generalized SBP property
\[
\bm{Q}^i = \bm{E}^T\bm{B}^i\bm{E} - \LRp{\bm{Q}^i}^T.
\]
Then, the decoupled SBP operator $\bm{D}^i_N$ (\ref{eq:QN}) satisfies a summation by parts property:
\begin{gather}
\bm{Q}^i_N+\LRp{\bm{Q}^i_N}^T = \bm{B}^i_N, \qquad \bm{B}^i_N = \LRp{\begin{array}{cc}\bm{0}&\\ & \bm{B}^i\end{array}}.\label{eq:dsbp}
\end{gather}
%\begin{enumerate}
%\item The mass matrix is positive definite, 
%\item The quadrature integrates exactly $\int_{\hat{D}} \pd{u}{x_i} v$ for all $u,v \in V^N\LRp{\hat{D}}$ and $i = 1,\ldots, d$.
%\end{enumerate}
\label{lemma:dsbp}
\end{theorem}
\begin{proof}
The proof is a straightforward extension of Theorem 1 in \cite{chan2017discretely} to polynomial approximation spaces on non-simplicial elements.  
\end{proof}

@@

\begin{remark}
It should be stressed that sufficiently accurate volume and surface quadratures are not necessary conditions for Theorem~\ref{lemma:dsbp}.  For example, it is well known that the use of $(N+1)$ point Gauss-Legendre-Lobatto (GLL) rules for both volume and surface quadratures result in a generalized SBP property, despite the fact that these rules are only accurate for degree $(2N-1)$ polynomials.  
%suppose $\hat{D}$ is a quadrilateral element, and that the volume quadrature is a tensor product of $(N+1)$ point one-dimensional Gauss-Legendre-Lobatto (GLL) rule.  If the surface quadrature also taken to be an $(N+1)$-point GLL rule over each face, then $\bm{Q}^i$ satisfies a traditional SBP property, which implies the decoupled SBP property for $\bm{Q}^i_N$.  More generally, it is straightforward to show that Theorem~\ref{lemma:dsbp} holds whenever $\bm{Q}^i$ satisfies a generalized SBP property 
%\[
%\bm{Q}^i = \bm{E}^T\bm{B}^i\bm{E}^T - \LRp{\bm{Q}^i}^T.
%\]
\label{remark:dsbp}
\end{remark}

@@

Throughout the remainder of this work, we make the following assumptions:
\begin{assumption}
\label{ass:quad}
Let $v \in V^{N}$ denote some fixed polynomial.  We assume that: 
\begin{enumerate}
\item the mass matrix $\bm{M}$ is positive definite under the volume quadrature rule,
\item the volume quadrature rule is exact for integrals of the form\\$\int_{\hat{D}} \pd{u}{\hat{x}_j} v$ for all $u \in V^N\LRp{\hat{D}}$, $j = 1,\ldots, d$.
\item the surface quadrature rule is exact for integrals of the form\\$\int_{\partial \hat{D}} u v \hat{n}_j$ for all $u \in V^N\LRp{\hat{D}}$, $j = 1,\ldots, d$, and $f \in \partial \hat{D}$.  
\end{enumerate}
\end{assumption}

@@

The conditions of Assumption~\ref{ass:quad} imply the following variational SBP property:
\begin{lemma}
\label{lemma:vsbp}
%Assume that $v \in V^M$.  
%If the volume quadrature is exact for polynomials of degree $N+M-1$ and the surface quadrature is exact for polynomials of degree $N+M$, 
Suppose Assumption~\ref{ass:quad} holds for $v(\bm{x})\in V^N$, and let $\bm{v}_q$ denote the vector of values of $v(\bm{x})$ of at volume quadrature points. Let $u(\bm{x})$ be an $L^2$ integrable function, and let $\bm{u}_q$ denote the values of $u(\bm{x})$ at quadrature points.  Then, 
\[
\bm{v}_q^T\bm{Q}_i \bm{u}_q = \bm{v}_q^T\LRp{ {\bm{E}}^T \bm{B}^i\bm{E} - \bm{Q}_i^T}\bm{u}_q.
\]
\end{lemma}
\begin{proof}
Recall that $\bm{Q}_i = \bm{W} \bm{V}_q \bm{D}^i\bm{P}_q$.  The quadrature-based $L^2$ projection $\Pi_Nu$ is a polynomial of degree $N$, and is computed discretely as $\bm{P}_q\bm{u}_q$.  By the exactness of the volume quadrature in Assumption~\ref{ass:quad}, 
\begin{align*}
\bm{v}_q^T\bm{Q}_i \bm{u}_q &= \bm{v}_q^T\bm{W} \bm{D}^i \bm{P}_q \bm{u}_q = \int_{\hat{D}} \pd{\Pi_N u}{\hat{x}_i} v = \int_{\partial \hat{D}} (\Pi_N u) v \hat{n}_i - \int_{\hat{D}} \LRp{\Pi_N u} \pd{v}{\hat{x}_i}.
\end{align*}
%Since the volume integrand $\LRp{\Pi_N u} \pd{v}{\hat{x}_i} \in V^{N+M-1}$, it is exactly integrated using the volume quadrature.  
%Additionally, since $\Pi_N u \in V^N$ and $v\in V^M$ on the surface of $\hat{D}$, the surface quadrature is exact for the surface integral and 
By the exactness of the surface quadrature in Assumption~\ref{ass:quad}, 
\begin{align*}
\int_{\partial \hat{D}} (\Pi_N u) v \hat{n}_i -   \int_{\hat{D}} \LRp{\Pi_N u} \pd{v}{\hat{x}_i} &=\bm{v}_f^T\bm{B}^i \bm{E}\bm{u}_q - \LRp{\bm{V}_q\bm{D}^i\bm{P}_q\bm{v}_q}^T\bm{W} \bm{V}_q\bm{P}_q\bm{u}_q\\
=& \bm{v}_q^T {\bm{E}}^T\bm{B}^i \bm{E}\bm{u}_q - \bm{v}_q^T \bm{P}_q^T\LRp{\bm{D}^i}^T \bm{V}_q^T \bm{W} \bm{V}_q\bm{P}_q\bm{u}_q,
\end{align*}
where we have used that, since $v \in V^N$, $\bm{v}_f = \bm{V}_f\bm{P}_q\bm{v}_q = \bm{E}\bm{v}_q$.  The proof is completed by noting that 
\[
\bm{P}_q^T\LRp{\bm{D}^i}^T \bm{V}_q^T \bm{W} \bm{V}_q\bm{P}_q = \bm{P}_q^T\LRp{\bm{D}^i}^T \bm{M}\bm{P}_q = \bm{P}_q^T\LRp{\bm{D}^i}^T \bm{V}_q^T\bm{W} = \bm{Q}_i^T.
\]
\end{proof}

>>>>

%\pdfoutput=1

\documentclass{svjour3}                     % onecolumn (standard format)
%\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
%\documentclass[smallextended]{svjour3}       % onecolumn (second format)
%\documentclass[twocolumn]{svjour3}          % twocolumn
%\usepackage[T1]{fontenc}

@@

\makeatletter
\renewcommand*{\@thmcounterend}{}
\makeatother
%\spnewtheorem{lemma}{Lemma}{\bfseries}{\itshape}
\spnewtheorem{assumption}{Assumption}{\bfseries}{\itshape}
%\spnewtheorem{remark}{Remark}{\bfseries}{\itshape}

%\usepackage{fullpage}
\usepackage{cite}

%\usepackage{slashbox} % for slashbox in tables

\usepackage{amsmath, amsfonts, amssymb}
%\newtheorem*{remark}{Remark}
%\theoremstyle{assumption}
%\newtheorem{assumption}{Assumption}
\usepackage{hyperref}
\usepackage[titletoc,toc,title]{appendix}
\usepackage{tikz}
%\usepackage{hhline}
\usepackage{graphicx, color, bm}
\usepackage{subfig}
\usepackage{pgfplots}
\usepackage{pgfplotstable}
\definecolor{markercolor}{RGB}{124.9, 255, 160.65}
%\usepackage{slopetri} % my own package for drawing triangle slopes

@@

\date{}
\author{Jesse Chan}
\title{Skew-symmetric entropy stable modal discontinuous Galerkin formulations}% with applications to hybrid meshes}
\titlerunning{Skew-symmetric entropy stable DG formulations}

@@

\begin{abstract}
High order entropy stable discontinuous Galerkin (DG) methods for nonlinear conservation laws satisfy an inherent discrete entropy inequality.  The construction of such schemes has relied on the use of carefully chosen nodal points \cite{gassner2013skew, fisher2013high, carpenter2014entropy, crean2018entropy, chan2018efficient} or volume and surface quadrature rules \cite{chan2017discretely, chan2018discretely} to produce operators which satisfy a summation-by-parts (SBP) property.  In this work, we show how to construct ``modal'' skew-symmetric DG formulations which are entropy stable for volume and surface quadratures under which a traditional SBP property does not hold.  These skew-symmetric formulations avoid the use of a ``strong'' matrix-based SBP property, and instead rely on the assumption that discrete operators exactly differentiate constants and satisfy a discrete form of the fundamental theorem of calculus.   We conclude with numerical experiments verifying the accuracy and stability of the proposed formulations, and discuss an application of skew-symmetric formulations for entropy stable DG schemes on mixed quadrilateral-triangle meshes.  
\end{abstract}

%\tableofcontents

@@

These issues have motivated the introduction of high order \textit{entropy stable} discretizations, which satisfy a semi-discrete entropy inequality while maintaining high order accuracy in smooth regions.  Proofs of continuous entropy inequalities rely on the chain rule, which does not hold discretely due to effects such as quadrature error.   Entropy stable schemes circumvent the loss of the chain rule while maintaining a semi-discrete analogue of the continuous entropy inequality by using an approach referred to as ``flux differencing''.  These schemes were first introduced as high order collocation methods on tensor product elements in \cite{fisher2013high, carpenter2014entropy, gassner2016split, gassner2017br1}, and were extended to simplicial elements in \cite{crean2017high, chen2017entropy, crean2018entropy, chan2017discretely, chan2018discretely}.  Entropy stable methods have also been extended to a variety of other discretization settings, including staggered grids \cite{parsani2016entropy}, Gauss-Legendre collocation \cite{chan2018efficient}, and non-conforming meshes \cite{friedrich2017entropy}.

Entropy stable ``modal'' DG discretizations \cite{chan2017discretely, chan2018discretely} are built upon flux differencing and a summation-by-parts (SBP) property, which serves as a discrete analogue of integration by parts for quadrature-based discretization matrices.  However, the SBP property does not hold for certain under-integrated quadrature rules, which arise naturally in some discretization settings.  For example, on hybrid meshes consisting of both quadrilateral and triangular elements, it is convenient to utilize the same quadrature rule on shared faces between different element types.  On degree $N$ quadrilateral elements, a popular choice of quadrature is an $(N+1)$-point Gauss-Legendre-Lobatto (GLL) rule.  When both volume and surface integrals are approximated using $(N+1)$ point GLL quadrature rules, the SBP property holds, despite the fact that GLL quadrature is inexact for the integrands which appear in finite element formulations \cite{fisher2013high}.  However, while GLL quadrature induces an SBP property on quadrilateral elements, it does not guarantee an SBP property if used on triangular elements \cite{chan2017discretely}.

@@

It should be noted that a similar approach to entropy stable discretizations was introduced within a finite difference framework \cite{chen2017entropy, crean2018entropy} using multidimensional differencing operators which satisfy accuracy conditions and an SBP property \cite{hicken2016multidimensional}.  These operators exist for nodal points corresponding to sufficiently accurate choices of volume and surface quadrature, but do not correspond to any specific basis or approximation space.  The formulations in \cite{chen2017entropy, crean2018entropy} differ from the ones presented in this work in that they are based on SBP finite differences and ``nodal'' (rather than ``modal'') DG formulations, with differentiation operators computed algebraically or through an optimization problem for each specific choice of nodes.  In contrast, ``modal'' formulations induce quadrature-based operators from an explicit approximation space, and accomodate general choices of volume and surface quadrature (e.g.\ volume quadratures without boundary nodes and over-integrated quadrature rules).

The structure of the paper is as follows: Section~\ref{sec:nonlin} describes the continuous entropy inequality which we aim to replicate discretely.  Section~\ref{sec:approx} and Section~\ref{sec:sbp} introduce polynomial approximation spaces and quadrature-based SBP operators on simplicial and tensor product elements.  Section~\ref{sec:skew1} then introduces the variational SBP property and an entropy conservative skew-symmetric formulation.  Section~\ref{sec:skew2} extends the skew-symmetric formulation to curved elements, and Section~\ref{sec:quadacc} provides explicit conditions for entropy stability in terms of quadrature accuracy and the polynomial degree of geometric mappings.  Section~\ref{sec:num} concludes by presenting numerical experiments which verify the theoretical assumptions, stability, and accuracy of the proposed formulations.

@@

The proof of (\ref{eq:entropyineq}) requires the use of the chain rule \cite{mock1980systems, harten1983symmetric, dafermos2005compensated}.  The instability-in-practice of high order schemes for (\ref{eq:nonlineqs}) can be attributed in part to the fact that the discrete form of the equations do not satisfy the chain rule, and thus do not satisfy (\ref{eq:entropyineq}).  As a result, discretizations of (\ref{eq:nonlineqs}) do not typically possess an underlying statement of stability.  This can be offset in practice by the numerical dissipation inherent in lower order schemes.  However, because high order discretizations possess low numerical dissipation, the lack of an underlying discrete stability has contributed to the perception that high order methods are inherently less stable than low order methods.

@@

In this work, we consider either simplicial reference elements (triangles and tetrahedra) or tensor product reference elements (quadrilaterals and hexahedra).  We define an approximation space using degree $N$ polynomials on the reference element; however, the natural polynomial approximation space differs depending on the element type \cite{chan2015gpu}.  On a $d$-dimensional reference simplex, the natural polynomial space consists of total degree $N$ polynomials 
\[
P^N\LRp{\hat{D}} = \LRc{\hat{x}_1^{i_1}\ldots\hat{x}_d^{i_d}, \quad \hat{\bm{x}} \in \hat{D}, \quad 0\leq \sum_{k=1}^d i_k \leq N}.
\]
In contrast, the natural polynomial space on a $d$-dimensional tensor product element is the space of maximum degree $N$ polynomials
\[
Q^N\LRp{\hat{D}} = \LRc{\hat{x}_1^{i_1}\ldots\hat{x}_d^{i_d}, \quad \hat{\bm{x}} \in \hat{D}, \quad 0\leq i_k \leq N, \quad k = 1,\ldots, d}.
\]
We denote the natural approximation space on a given reference element $\hat{D}$ by $V^N$.  Furthermore, we denote the dimension of $V^N$ as $N_p = {\rm dim}\LRp{V^N\LRp{\hat{D}}}$.

The proofs presented in this work will also refer to anisotropic tensor product polynomial spaces, where the maximum polynomial degree varies depending on the coordinate direction.  We denote such spaces by $Q^{N_1, \ldots, N_d}$, where $N_k$ are non-negative integers and
\[
Q^{N_1, N_2, \ldots, N_d}\LRp{\hat{D}} = \LRc{\hat{x}_1^{i_1}\ldots\hat{x}_d^{i_d}, \quad \hat{\bm{x}} \in \hat{D}, \quad 0\leq i_k \leq N_k, \quad k = 1,\ldots, d}.
\]
For example, the isotropic tensor product space $Q^N$ is the same as $Q^{N,\ldots,N}$.

@@

It was shown in \cite{chan2017discretely, chan2018efficient} that, when combined with projection and lifting matrices, $\bm{D}^i_N$ produces a high order polynomial approximation of $f\pd{g}{x_i}$. 
Let $f, g$ be differentiable functions, and let $\bm{f}_i = f(\bm{x}_i)$, $\bm{g}_i = g(\bm{x}_i)$ denote values of $f,g$ at volume and surface quadrature points.  Then, $f\pd{g}{x_i}$ can be approximated by the polynomial with modal coefficients given by
\begin{gather}
f\pd{g}{x_i} \approx \LRs{\begin{array}{cc}
\bm{P}_q & \bm{L}_f\end{array}} {\rm diag}\LRp{\bm{f}}\bm{D}^i_N \bm{g}.  
\label{eq:dnapprox}
\end{gather}
%The approximation can also be interpreted as solving a variational problem.  Let $n_i$ be the $i$th component of the unit normal on $\hat{D}$.  Then (\ref{eq:dnapprox}) is equivalent to finding $u(\bm{x}) \approx f\pd{g}{x_i}$ such that, for all $v\in V^N\LRp{\hat{D}}$, 
%\begin{align*}
%\int_{\hat{D}} u(\bm{x})v(\bm{x}) = \int_{\hat{D}}{g\pd{\Pi_Nf}{x_i}v} + \int_{\partial \hat{D}}{(f-\Pi_Nf)\frac{\LRp{gv + \Pi_N(gv)}}{2}}\hat{n}_i\hat{J}_f.
%\end{align*}

For sufficiently accurate quadrature rules, the matrix $\bm{Q}^i_N$ also satisfies a decoupled summation-by-parts (SBP) property, which is used to prove semi-discrete entropy stability for nonlinear conservation laws.  
\begin{theorem}%[Theorem 1 in \cite{chan2017discretely}]
If $\bm{Q}^i$ satisfies a generalized SBP property
\[
\bm{Q}^i = \bm{E}^T\bm{B}^i\bm{E} - \LRp{\bm{Q}^i}^T,
\]
then the decoupled SBP operator $\bm{D}^i_N$ (\ref{eq:QN}) satisfies a summation by parts property:
\begin{gather}
\bm{Q}^i_N+\LRp{\bm{Q}^i_N}^T = \bm{B}^i_N, \qquad \bm{B}^i_N = \LRp{\begin{array}{cc}\bm{0}&\\ & \bm{B}^i\end{array}}.\label{eq:dsbp}
\end{gather}
%\begin{enumerate}
%\item The mass matrix is positive definite, 
%\item The quadrature integrates exactly $\int_{\hat{D}} \pd{u}{x_i} v$ for all $u,v \in V^N\LRp{\hat{D}}$ and $i = 1,\ldots, d$.
%\end{enumerate}
\label{lemma:dsbp}
\end{theorem}
\begin{proof}
The proof is a straightforward extension of Theorem 1 in \cite{chan2017discretely} to polynomial approximation spaces on non-simplicial elements.  
\qed\end{proof}

@@

\begin{remark}
It should be stressed that the accuracy conditions on volume and surface quadratures are sufficient but not necessary conditions for Theorem~\ref{lemma:dsbp}.  For example, it is well known that the use of $(N+1)$ point Gauss-Legendre-Lobatto (GLL) rules for both volume and surface quadratures result in a generalized SBP property, despite the fact that these rules are only accurate for degree $(2N-1)$ polynomials.  
%suppose $\hat{D}$ is a quadrilateral element, and that the volume quadrature is a tensor product of $(N+1)$ point one-dimensional Gauss-Legendre-Lobatto (GLL) rule.  If the surface quadrature also taken to be an $(N+1)$-point GLL rule over each face, then $\bm{Q}^i$ satisfies a traditional SBP property, which implies the decoupled SBP property for $\bm{Q}^i_N$.  More generally, it is straightforward to show that Theorem~\ref{lemma:dsbp} holds whenever $\bm{Q}^i$ satisfies a generalized SBP property 
%\[
%\bm{Q}^i = \bm{E}^T\bm{B}^i\bm{E}^T - \LRp{\bm{Q}^i}^T.
%\]
\label{remark:dsbp}
\end{remark}

@@

Throughout the remainder of this work, we will make the following assumption for specific functions $v(\bm{x})$:
\begin{assumption}
\label{ass:quad}
Let $v \in V^{N}$ denote some fixed polynomial.  We assume that: 
\begin{enumerate}
\item the mass matrix $\bm{M}$ is positive definite under the volume quadrature rule,
\item the volume quadrature rule is exact for integrals of the form\\$\int_{\hat{D}} \pd{u}{\hat{x}_j} v$ for all $u \in V^N\LRp{\hat{D}}$, $j = 1,\ldots, d$.
\item the surface quadrature rule is exact for integrals of the form\\$\int_{\partial \hat{D}} u v \hat{n}_j$ for all $u \in V^N\LRp{\hat{D}}$, $j = 1,\ldots, d$, and $f \in \partial \hat{D}$.  
\end{enumerate}
\end{assumption}

@@

The conditions of Assumption~\ref{ass:quad} imply the following variational SBP property:
\begin{lemma}
\label{lemma:vsbp}
%Assume that $v \in V^M$.  
%If the volume quadrature is exact for polynomials of degree $N+M-1$ and the surface quadrature is exact for polynomials of degree $N+M$, 
Suppose Assumption~\ref{ass:quad} holds for $v(\bm{x})\in V^N$, and let $\bm{v}_q$ denote the vector of values of $v(\bm{x})$ of at volume quadrature points. Let $u(\bm{x})$ be an $L^2$ integrable function, and let $\bm{u}_q$ denote the values of $u(\bm{x})$ at quadrature points.  Then, 
\[
\bm{v}_q^T\bm{Q}_i \bm{u}_q = \bm{v}_q^T\LRp{ {\bm{E}}^T \bm{B}^i\bm{E} - \bm{Q}_i^T}\bm{u}_q.
\]
\end{lemma}
\begin{proof}
Recall that $\bm{Q}_i = \bm{W} \bm{V}_q \bm{D}^i\bm{P}_q$.  The quadrature-based $L^2$ projection $\Pi_Nu$ is a polynomial of degree $N$, and is computed discretely as $\bm{P}_q\bm{u}_q$.  By the exactness of the volume quadrature in Assumption~\ref{ass:quad}, 
\begin{align*}
\bm{v}_q^T\bm{Q}_i \bm{u}_q &= \bm{v}_q^T\bm{W} \bm{D}^i \bm{P}_q \bm{u}_q = \int_{\hat{D}} \pd{\Pi_N u}{\hat{x}_i} v = \int_{\partial \hat{D}} (\Pi_N u) v \hat{n}_i - \int_{\hat{D}} \LRp{\Pi_N u} \pd{v}{\hat{x}_i}.
\end{align*}
%Since the volume integrand $\LRp{\Pi_N u} \pd{v}{\hat{x}_i} \in V^{N+M-1}$, it is exactly integrated using the volume quadrature.  
%Additionally, since $\Pi_N u \in V^N$ and $v\in V^M$ on the surface of $\hat{D}$, the surface quadrature is exact for the surface integral and 
By the exactness of the surface quadrature in Assumption~\ref{ass:quad}, 
\begin{align*}
\int_{\partial \hat{D}} (\Pi_N u) v \hat{n}_i -   \int_{\hat{D}} \LRp{\Pi_N u} \pd{v}{\hat{x}_i} &=\bm{v}_f^T\bm{B}^i \bm{E}\bm{u}_q - \LRp{\bm{V}_q\bm{D}^i\bm{P}_q\bm{v}_q}^T\bm{W} \bm{V}_q\bm{P}_q\bm{u}_q\\
=& \bm{v}_q^T {\bm{E}}^T\bm{B}^i \bm{E}\bm{u}_q - \bm{v}_q^T \bm{P}_q^T\LRp{\bm{D}^i}^T \bm{V}_q^T \bm{W} \bm{V}_q\bm{P}_q\bm{u}_q,
\end{align*}
where we have used that, since $v \in V^N$, $\bm{v}_f = \bm{V}_f\bm{P}_q\bm{v}_q = \bm{E}\bm{v}_q$.  The proof is completed by noting that 
\[
\bm{P}_q^T\LRp{\bm{D}^i}^T \bm{V}_q^T \bm{W} \bm{V}_q\bm{P}_q = \bm{P}_q^T\LRp{\bm{D}^i}^T \bm{M}\bm{P}_q = \bm{P}_q^T\LRp{\bm{D}^i}^T \bm{V}_q^T\bm{W} = \bm{Q}_i^T.
\]
\qed\end{proof}
